---
title: "Response of Guinea Pig Incisor Teeth Odontoblasts to Vitamin C Dose and Delivery Method"
author: "Jonathan Owen"
date: "January 31, 2016"
output: rmarkdown::tufte_handout
---

Introduction
============
Statistical inference on whether variations of guinea pig tooth length were 
related to the dose and form of vitamin C supplmentation in their diets is 
reported here.

Data
===
Data are taken from a monograph by C.I. Bliss[^1] but were originally reported 
by E.W. Crampton.[^2]  A total of 60 guinea pigs were randomly assigned to 
treatment groups of 10 that were fed a basal diet and supplemental vitamin C as 
either orange juice or pure ascorbic acid at one of three dose levels 
(0.5, 1.0, and 2.0 mg per day).  Measurements on the incisor teeth of each 
guinea pig were made after 42 days of feeding. The response is the length of 
odontoblast[^3] cells observed under a microscope (at 440x magnification).  
\newthought{In the data file} there are 60 observations of 3 variables  
`len` \enspace \enspace odontoblast length, a numeric value  
`supp` \enspace  vitamin C delivery method, a 2-level factor with values  
 \ \enspace \enspace \enspace \enspace \enspace ascorbic acid (`VC`) or orange 
 juice (`OJ`)  
`dose` \enspace  vitamin C dose in mg per day with values 0.5, 1.0, and 2.0




Exploratory Data Analysis
==========================
A basic summary of the `ToothGrowth` data is included in Table 1 and, in Figure 
1, a scatterplot shows the variation of odontoblast length with vitamin C dose. In the 
plot, delivery method is indicated by the color of the points. The mean 
odontoblast lengths from Table 1 are included next to the relevant data in 
Figure 1. 

```{r, meansTooth, echo = FALSE, message = FALSE, results = "asis"}
library(dplyr)
library(ztable)
library(xtable)
options(xtable.type = "latex")

edaTooth <- ToothGrowth %>%
              group_by(supp, dose) %>%
              summarise(mLen = mean(len), vLen = sd(len)^2)
edaTooth <- data.frame(edaTooth)
oldNames <- names(edaTooth)
prepXTooth <- cbind(c("ascorbic acid", "", "", "orange juice", "", ""),edaTooth)
names(prepXTooth) <- c("supplement", "", "dose", "mean", "variance")
xTooth <- xtable(prepXTooth[, -2],
                 caption = "Summary of \\texttt{ToothGrowth} data")
print(xTooth, 
      include.rownames = FALSE, 
      comment = FALSE)


```  



\noindent The summarized data and plot suggest that odontoblast length increases 
with vitamin C dose. The two delivery methods give different response 
levels as odontoblasts in guinea pigs that were given orange juice (red points) 
appear to be longer than those given an equivalent dose of ascorbic acid (blue 
points). An exception to this might be for the 2.0 mg per day dose, for which 
the mean odontoblast length is similar for vitamin C delivered as 
orange juice (26.06) and ascorbic acid (26.14). Related to this, for orange 
juice the increase in odontoblast length between 1.0 mg per day (22.70) and 2.0 
mg per day (26.06) may not be large enough to be significant.




\newthought {Density curves} are included in the right margin of the scatterplot. 
These approximate the distributions of odontoblast lengths with a moving average 
of frequency. The implications of the shapes of these curves for assumptions in 
hypothesis testing will be discussed below.

```{r plotcode, echo=FALSE, message = FALSE, fig.cap = "Scatterplot (with jitter added) of guinea pig incisor odontoblast length vs vitamin C dose delivered as either pure ascorbic acid or orange juice. Marginal density plots show the distribution of osteoblast lengths for each treatment."}
library(ggplot2)
library(ggthemes)
library(ggExtra)
library(grid)
library(gridExtra)
library(htmlTable)

#names(meansTooth) <- oldNames

jitterTooth <- ggplot(data = ToothGrowth,
                      aes(x = dose, y = len, 
                          color = factor(supp, labels =c("orange juice", 
                                                         "ascorbic acid"))))   +
             theme_bw()                                                        +
             theme(text = element_text(size = 7, family = "serif"),
                   legend.title = element_blank(),
                   legend.key = element_blank(),
                   legend.key.height = unit(0.5, "line"),
                   legend.key.width = unit(0.5, "line"),
                   legend.background = element_blank(),
                   legend.position = c(0.175, 0.95))                           +
             geom_jitter(shape = 19, alpha = 0.5,
                         position = position_jitter(width = 0.05))             +
             labs(x = "dose, mg vitamin C per day", y = "odontoblast length")  

densityTooth <- ggplot()                                                       + 
                facet_grid(~ dose)                                             +
                theme(text = element_text(size = 7, family = "serif"),
                      legend.position = "none",
                      panel.background = element_blank(),
                      panel.grid.major = element_blank(),
                      panel.grid.minor = element_blank(),
                      axis.text = element_blank(),
                      axis.ticks = element_blank(),
                      axis.title.x = element_text(hjust = 0.35),
                      axis.title.y = element_blank(),
                      strip.text.x = element_blank(),
                      strip.background = element_blank(),
                      plot.margin = unit(c(0.4, 0, 0.5, -1.5), "lines"))       +
                geom_line(data = ToothGrowth,
                          aes(x = len, 
                              color = factor(supp)),
                          alpha = 0.75,
                          stat = "density")                                    +
                geom_text(data = edaTooth, 
                          aes(x = mLen, 
                              y = c(0.17, 0.17, 0.16, 0.16, 0.15, 0.07), 
                              label = format(mLen, digits = 4),  
                              color = factor(supp)),
                          size = 1.75, family = "serif", 
                          hjust = 1, vjust = 1)                               +
                coord_flip()

grid.arrange(jitterTooth, densityTooth, ncol = 2, widths = c(4,2.5))
```  


Hypothesis Tests
================
Student's t-tests were used to test multiple hypotheses of whether 
tooth lengths for two treatments were significantly different. For each t-test, 
it is assumed that the data are sampled from independent and identically 
distributed populations.
The density curves in Figure 1 address the validity of the assumption.  Although 
curves are not exactly those of a normal distribution, they are consistent with 
what might be observed for relatively small samples (n = 10) from a normal 
distribution.
Based on the variances in Table 1 and the width of the density curves in Figure 
1, it did not seem appropriate to assume that the populations have equal variance.
Testing multiple hypotheses can introduce errors. Using $\alpha$ = 0.05 means we 
expect that in 1 out of every 20 tests, the null hypothesis would incorrectly be rejected 
(type 1 error). To control this type of error, adjusted $\alpha$ values were used. The 
adjustment is based on the rank of the calculated $p$-value. For *i*^th^ lowest $p$-value of the 9 tests, $\alpha_i$ is 
taken as its rank multiplied by 0.5/9 (approximately 0.0056). If $p_i$ is less than $\alpha_i$ then the null hypothesis 
is rejected.  The hypothesis tests with $p_i$ and $\alpha_i$ are reported in Table 2.  


```{r hypothesistest, echo = FALSE, results="asis", message = FALSE} 
dfH <- data.frame(supp1  = c(rep("VC", 3), rep("OJ", 3), rep("VC", 3)), 
                            dose1  = rep(c(0.5, 1.0, 2.0), 3),
                            supp2  = c(rep("VC", 3), rep("OJ", 6)),
                            dose2  = c(rep(c(1.0, 2.0, 0.5), 2),
                                       0.5, 1.0, 2.0))



for (i in 1:9)  {
  outTests <- t.test(ToothGrowth[ToothGrowth$supp == dfH[i, 1] & 
                                         ToothGrowth$dose == dfH[i, 2],1],
                             ToothGrowth[ToothGrowth$supp == dfH[i, 3] & 
                                         ToothGrowth$dose == dfH[i, 4],1],
                     paired = FALSE,
                     var.equal = FALSE)
  dfH$diffM[i]   <- outTests$estimate[1] - outTests$estimate[2]
  dfH$lowCI[i]   <- outTests$conf.int[1]
  dfH$upCI[i]  <- outTests$conf.int[2]
  dfH$p[i]  <- outTests$p.value
}
dfH$rankP <- rank(dfH$p)
dfH$iAlpha <- dfH$rankP * 0.05/9
dfH <- cbind(dfH, c("ascorbic acid", "", "", "orange juice", "", "", "ascorbic acid", "vs", "orange juice"))
names(dfH) <- c("supp1", " ", "supp2", "  ", "$\\Delta_{mean}$", "   ", "    ", "     ", "rank", "      ", "")


xH <- xtable(dfH[,c(11, 2, 4, 6, 5, 7:8, 10)],
             digits = c(0, 0, 1, 1, 2, 2, 2, 4, 4),
             align = c(rep("l", 2), rep("r", 5), rep("c", 2)),
             caption = "Hypothesis tests on \\texttt{ToothGrowth} data set. Different doses (in mg per day) are compared for each delivery method. Then the two delivery methods are compared for each dose. Unadjusted $p$ values and an $\\alpha$ value adjusted by the $p$ value rank are reported for each test.")
addtorow <- list()
addtorow$pos <- list(0)
addtorow$command <- "&  \\multicolumn{2}{c}{dose} &{2.5\\%} &{50\\%} &{97.5\\%} &{$p$}&{$\\alpha_i$}\\\\\n"

print(xH, 
      include.rownames = FALSE, 
      comment = FALSE,
      add.to.row = addtorow,
      sanitize.text.function = function(x) {x})                

```  

Conclusions
===========
Based on the assumptions described in the *Hypothesis Tests* section and using 
corrections for testing multiple hypotheses, the following conclusions can be made. 
The null hypotheses should be rejected for all tests apart from comparing 2 mg/day ascorbic 
acid to 2 mg/day orange juice. This means that  
(i) Odontoblasts lengths differ significantly between any pair of doses as ascorbic acid, with the length increasing as 
dose increases  
(ii) Odontoblasts lengths differ significantly between any pair of doses orange juice, with the length increasing as 
dose increases  
(iii) Odontoblasts lengths differ significantly between ascorbic acid and orange juice at 0.5 mg/day, with the length being greater 
when delivery is by orange juice  
(iv) Odontoblasts lengths differ significantly between ascorbic acid and orange juice at 1.0 mg/day, with the length being greater 
when delivery is by orange juice  
  
\newpage

Appendix
========
```{r, ref.label = "plotcode", eval=FALSE}
```  

```{r, ref.label = "hypothesistest", eval=FALSE}
```  
  
[^1]:  C.I. Bliss.  
The Statistics of Bioassay.  
Academic Press, New York, 1952.  

[^2]:  E.W. Crampton.  
The Growth of the Odontoblasts of  
the Incisor Tooth as a Criterion of the  
Vitamin C Intake of the Guinea Pig.  
*J. Nutrition* **33** 491-504 (1947).  

[^3]:  Odontoblasts are columnar cells  
lying at the interface of a tooth's inner  
enamel surface with the dental pulp.  
Their function is associated with the  
production of dentine.  

